---
title: "Your IEEE Paper Title Here"
classoption: [onecolumn]  # <--- Add this line here
format:
  ieee-pdf:
    keep-tex: true  # Optional: keeps the raw .tex file for debugging
execute:
  cache: true
author:
  - name: "Your Name"
    affiliations:
      - name: "Pop!_OS University"
        department: "Department of Data Science"
        city: "Pendleton"
        country: "USA"
    email: "user@example.com"
    attributes:
      corresponding: true
  - name: "Co-Author Name"
    affiliations:
      - name: "Another Institute"
abstract: |
  This is the abstract of your paper. It should be a concise summary of your research, methodology, and results.
keywords: [Quarto, RStudio, IEEE, Linux]
bibliography: bibliography.bib
cite-method: natbib
---


```{r}
library(data.table)
library(tidyverse)
library(pheatmap)
library(lsa) 
setDTthreads(0) 

# --- METRIC CONFIGURATION FLAGS ---
run_jaccard    <- TRUE
run_tfidf      <- TRUE
run_overlap    <- TRUE
run_dice       <- TRUE
run_kulczynski <- TRUE
run_pmi        <- TRUE
run_maxscale   <- TRUE

```

Executive Summary
This document expands the similarity analysis between majors by introducing advanced metrics. While Jaccard and Cosine are standard, they often over-represent "Gen-Ed" overlap (e.g., almost every STEM student taking Calculus I).By adding TF-IDF and Overlap Coefficients, we can better isolate the unique curricular bonds between majors.

# Data Loading and Preparation

```{r}
degree_path <- "data/degree_2020_03_16_fix9a.csv"
course_path <- "data/course_2020_03_16_fix9a.csv"

degree_dt <- fread(degree_path, select = c("mcid", "cip6", "degree"))
# Pad to 6 digits to preserve leading zeros (e.g., CIP 04.0401 → "040401" not "40401")
degree_dt[, cip6 := str_pad(as.character(cip6), width = 6, side = "left", pad = "0")]

course_dt <- fread(course_path, select = c("mcid", "abbrev", "number"))

# --- HEURISTICS: COURSE NAME NORMALIZATION ---
# 1. Standardization: Uppercase and Trim
course_dt[, abbrev := str_to_upper(str_trim(abbrev))]
course_dt[, number := str_to_upper(str_trim(number))]

# 2. Leading Zeros: Remove '0' from start of numbers (e.g., '0101' -> '101')
course_dt[, number := str_remove(number, "^0+")]

# 3. Suffix Stripping: Remove trailing single letters (e.g., '101H' -> '101', '400W' -> '400')
# This groups Honors (H), Writing (W), or Sequence (A/B) variants with the parent course.
course_dt[, number_clean := str_remove(number, "[A-Z]$")]

# 4. Construct Unified ID
course_dt[, course_id := paste(abbrev, number_clean)]
# ---------------------------------------------

valid_mcids <- intersect(degree_dt$mcid, course_dt$mcid)
degree_dt <- degree_dt[mcid %in% valid_mcids]
course_dt <- course_dt[mcid %in% valid_mcids]


```

# Vectorization and Aggregation

```{r}
combined_dt <- merge(degree_dt[, .(mcid, cip6)], 
                    course_dt[, .(mcid, course_id)], 
                    by = "mcid")

# Count total graduates per major to use for normalization
major_sizes <- degree_dt[, .(total_grads = uniqueN(mcid)), by = cip6]

# Count frequency of courses per major
major_course_weights <- combined_dt[, .(student_count = .N), by = .(cip6, course_id)]

# Add probability (percentage of students in major who took course) to normalize for major size
major_course_weights <- merge(major_course_weights, major_sizes, by = "cip6")
major_course_weights[, prob_taken := student_count / total_grads]

# Max-scaling: weight relative to the most popular course in THAT major
major_course_weights[, max_scaled := student_count / max(student_count), by = cip6]

# Unique list of courses per major
major_course_sets <- combined_dt[, .(course_set = list(unique(course_id))), by = cip6]

# Ensure 1:1 CIP-to-Degree mapping for labels
degree_names_map <- unique(degree_dt[, .(cip6, degree)])
degree_names_map <- degree_names_map[, .(degree = first(degree)), by = cip6]
setkey(degree_names_map, cip6)

rm(combined_dt)
gc()

```


# Advanced Metrics Calculation

```{r}
# --- CIP Group Definitions ---
cips_to_compare <- c(
  "110101", "110701", # Computing
  "140901", "140903", # Computer Eng
  "400601", "143901", # Geo/Geol Eng
  "260101", "260202", # Bio/Biochem
  "141901", "141001", # Mech/Elec Eng
  "270101", "400801"  # Math/Physics
)

# 1. TF-IDF Cosine
tfidf_sim <- NULL
if (run_tfidf) {
  subset_weights <- major_course_weights[cip6 %in% cips_to_compare]
  dtm <- dcast(subset_weights, cip6 ~ course_id, value.var = "prob_taken", fill = 0)
  cip_labels_cos <- as.character(dtm$cip6)
  dtm_matrix <- as.matrix(dtm[, -1])
  tf <- dtm_matrix 
  idf <- log(nrow(dtm_matrix) / (colSums(dtm_matrix > 0) + 1))
  tfidf_matrix <- t(t(tf) * idf)
  # Multiply by 100 to convert to percentage
  tfidf_sim <- lsa::cosine(t(tfidf_matrix)) * 100
  
  labels_tfidf <- degree_names_map[cip_labels_cos, degree]
  if(any(is.na(labels_tfidf))) labels_tfidf[is.na(labels_tfidf)] <- cip_labels_cos[is.na(labels_tfidf)]
  colnames(tfidf_sim) <- labels_tfidf; rownames(tfidf_sim) <- labels_tfidf
}

# 2. Max-Scaled Cosine
maxscale_sim <- NULL
if (run_maxscale) {
  subset_weights <- major_course_weights[cip6 %in% cips_to_compare]
  dtm_max <- dcast(subset_weights, cip6 ~ course_id, value.var = "max_scaled", fill = 0)
  cip_labels_max <- as.character(dtm_max$cip6)
  # Multiply by 100 to convert to percentage
  maxscale_sim <- lsa::cosine(t(as.matrix(dtm_max[, -1]))) * 100
  
  labels_max <- degree_names_map[cip_labels_max, degree]
  colnames(maxscale_sim) <- labels_max; rownames(maxscale_sim) <- labels_max
}

# 3. PMI (Pointwise Mutual Information) Similarity
pmi_sim <- NULL
if (run_pmi) {
  subset_weights <- major_course_weights[cip6 %in% cips_to_compare]
  total_n <- sum(subset_weights$student_count)
  p_major <- subset_weights[, .(p_m = sum(student_count)/total_n), by = cip6]
  p_course <- subset_weights[, .(p_c = sum(student_count)/total_n), by = course_id]
  
  pmi_data <- merge(subset_weights, p_major, by = "cip6")
  pmi_data <- merge(pmi_data, p_course, by = "course_id")
  pmi_data[, pmi := log2((student_count/total_n) / (p_m * p_c))]
  pmi_data[pmi < 0, pmi := 0] # Use Positive PMI
  
  dtm_pmi <- dcast(pmi_data, cip6 ~ course_id, value.var = "pmi", fill = 0)
  cip_labels_pmi <- as.character(dtm_pmi$cip6)
  # Multiply by 100 to convert to percentage
  pmi_sim <- lsa::cosine(t(as.matrix(dtm_pmi[, -1]))) * 100
  
  labels_pmi <- degree_names_map[cip_labels_pmi, degree]
  colnames(pmi_sim) <- labels_pmi; rownames(pmi_sim) <- labels_pmi
}

# 4. Binary Set Metrics
metrics_results <- NULL
if (run_jaccard | run_overlap | run_dice | run_kulczynski) {
  subset_sets <- major_course_sets[cip6 %in% cips_to_compare]
  metrics_results <- CJ(cip_a = cips_to_compare, cip_b = cips_to_compare)

  calc_metrics <- function(c1, c2) {
    set_a <- unlist(subset_sets[cip6 == c1, course_set]); set_b <- unlist(subset_sets[cip6 == c2, course_set])
    if(length(set_a) == 0 | length(set_b) == 0) return(list(jaccard = 0, overlap = 0, dice = 0, kulczynski = 0))
    inter <- length(intersect(set_a, set_b)); n_a <- length(set_a); n_b <- length(set_b)
    # Binary metrics multiplied by 100
    return(list(jaccard = 100 * (inter / length(union(set_a, set_b))), 
                overlap = 100 * (inter / min(n_a, n_b)),
                dice = 100 * ((2 * inter) / (n_a + n_b)), 
                kulczynski = 100 * (0.5 * ((inter / n_a) + (inter / n_b)))))
  }

  m_res <- mapply(calc_metrics, metrics_results$cip_a, metrics_results$cip_b, SIMPLIFY = FALSE)
  metrics_results[, jaccard := sapply(m_res, `[[`, "jaccard")]
  metrics_results[, overlap := sapply(m_res, `[[`, "overlap")]
  metrics_results[, dice := sapply(m_res, `[[`, "dice")]
  metrics_results[, kulczynski := sapply(m_res, `[[`, "kulczynski")]
}

```


# Metric Comparison

```{r}
clean_lab <- function(x) str_wrap(str_remove(x, "Bachelor of Science in "), 25)

format_mat <- function(res_dt, var) {
  mat <- dcast(res_dt, cip_a ~ cip_b, value.var = var)
  current_cips <- as.character(mat$cip_a); labels <- degree_names_map[current_cips, degree]
  if(any(is.na(labels))) labels[is.na(labels)] <- current_cips[is.na(labels)]
  m <- as.matrix(mat[, -1]); rownames(m) <- labels; colnames(m) <- labels
  return(m)
}

if (run_jaccard && !is.null(metrics_results)) {
  jac_m <- format_mat(metrics_results, "jaccard")
  pheatmap(jac_m, labels_row = clean_lab(rownames(jac_m)), labels_col = clean_lab(colnames(jac_m)),
           display_numbers = TRUE, number_format = "%.0f", main = "Jaccard Index (Binary Overlap)", fontsize = 9,
           color = colorRampPalette(c("white", "#feedde", "#fd8d3c"))(100))
}

if (run_tfidf && !is.null(tfidf_sim)) {
  pheatmap(tfidf_sim, labels_row = clean_lab(rownames(tfidf_sim)), labels_col = clean_lab(colnames(tfidf_sim)),
           display_numbers = TRUE, number_format = "%.0f", main = "TF-IDF Cosine (Specific Curricular Bonds)", fontsize = 9,
           color = colorRampPalette(c("white", "#f7fbff", "#08306b"))(100))
}

if (run_maxscale && !is.null(maxscale_sim)) {
  pheatmap(maxscale_sim, labels_row = clean_lab(rownames(maxscale_sim)), labels_col = clean_lab(colnames(maxscale_sim)),
           display_numbers = TRUE, number_format = "%.0f", main = "Max-Scaled Cosine (Internal Core)", fontsize = 9,
           color = colorRampPalette(c("white", "#f7fcf0", "#4eb3d3"))(100))
}

if (run_pmi && !is.null(pmi_sim)) {
  pheatmap(pmi_sim, labels_row = clean_lab(rownames(pmi_sim)), labels_col = clean_lab(colnames(pmi_sim)),
           display_numbers = TRUE, number_format = "%.0f", main = "PMI Similarity (Signature Bonds)", fontsize = 9,
           color = colorRampPalette(c("white", "#fff7bc", "#d95f0e"))(100))
}

if (run_overlap && !is.null(metrics_results)) {
  ovl_m <- format_mat(metrics_results, "overlap")
  pheatmap(ovl_m, labels_row = clean_lab(rownames(ovl_m)), labels_col = clean_lab(colnames(ovl_m)),
           display_numbers = TRUE, number_format = "%.0f", main = "Overlap Coefficient (Subset Detection)", fontsize = 9,
           color = colorRampPalette(c("white", "#e5f5e0", "#31a354"))(100))
}

if (run_dice && !is.null(metrics_results)) {
  dice_m <- format_mat(metrics_results, "dice")
  pheatmap(dice_m, labels_row = clean_lab(rownames(dice_m)), labels_col = clean_lab(colnames(dice_m)),
           display_numbers = TRUE, number_format = "%.0f", main = "Dice Similarity (Intersection Weighting)", fontsize = 9,
           color = colorRampPalette(c("white", "#f2f0f7", "#6a51a3"))(100))
}

if (run_kulczynski && !is.null(metrics_results)) {
  kul_m <- format_mat(metrics_results, "kulczynski")
  pheatmap(kul_m, labels_row = clean_lab(rownames(kul_m)), labels_col = clean_lab(colnames(kul_m)),
           display_numbers = TRUE, number_format = "%.0f", main = "Kulczynski Similarity (Mutual Resemblance)", fontsize = 9,
           color = colorRampPalette(c("white", "#fff5f0", "#cb181d"))(100))
}


```
# Curricular Similarity Metrics

## 1. Jaccard Index (Binary Overlap)
- **Logic**: The size of the intersection divided by the size of the union.
  $$J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$
- **Reflection on Data**: This is the baseline "breadth" metric. In the MIDFIELD dataset, it reflects the total shared curriculum. However, because it treats every course taken by even one student as a set member, it can be easily "diluted" by rare electives, often resulting in lower scores for very large or diverse departments.

## 2. TF-IDF Cosine (Specific Curricular Bonds)
- **Logic**: Weights courses based on their "specificity." A course like *Calculus I* (taken by many majors) gets low weight, while *Advanced Microcontrollers* (unique to a few) gets high weight.
  $$\text{Cosine Similarity} = \frac{\mathbf{v}_A \cdot \mathbf{v}_B}{\|\mathbf{v}_A\| \|\mathbf{v}_B\|}$$
- **Reflection on Data**: This metric effectively filters out the "STEM background noise." It highlights the specialized technical requirements that truly define a major's identity. It is highly resistant to major size because it uses normalized frequencies.

## 3. Overlap Coefficient (Subset Detection)
- **Logic**: Intersection size divided by the size of the smaller set.
  $$\text{Overlap}(A, B) = \frac{|A \cap B|}{\min(|A|, |B|)}$$
- **Reflection on Data**: Excellent for identifying "nested" curricula. For example, if Software Engineering students take almost everything Computer Science students take, but the CS students have a much broader elective set, the Overlap Coefficient will stay high while Jaccard drops. It reveals if one major is effectively a component of another.

## 4. PMI Similarity (Signature Bonds)
- **Logic**: Pointwise Mutual Information measures if a course-major pairing occurs more often than would be expected if they were independent.
  $$\text{PMI}(m, c) = \log \left( \frac{p(m, c)}{p(m)p(c)} \right)$$
- **Reflection on Data**: This is the most sensitive metric for "signature" behaviors. It reflects the courses that students are surprised to be in unless they are in that specific major. It is great for distinguishing between two very similar-sounding engineering programs.

## 5. Max-Scaled Cosine (Internal Core)
- **Logic**: Normalizes all course counts relative to the single most popular course within that major.
  $$x_{scaled} = \frac{x}{\max(\mathbf{x}_{major})}$$
- **Reflection on Data**: This reflects the "standard experience" of a major. By scaling everything to the major's own "anchor" course, it removes the effect of different graduation rates and focuses purely on the internal structure of what a "typical" student takes.

## 6. Dice Similarity (Intersection Weighting)
- **Logic**: Twice the intersection divided by the sum of the set sizes.
  $$\text{Dice}(A, B) = \frac{2 |A \cap B|}{|A| + |B|}$$
- **Reflection on Data**: Similar to Jaccard but gives more credit to the shared courses. It is less sensitive to the total variety of electives (the union) and more focused on the core intersection.

## 7. Kulczynski Similarity (Mutual Resemblance)
- **Logic**: The average of the two inclusion proportions (Intersection/A and Intersection/B).
  $$\text{Kulczynski}(A, B) = \frac{1}{2} \left( \frac{|A \cap B|}{|A|} + \frac{|A \cap B|}{|B|} \right)$$
- **Reflection on Data**: Reflects the average confidence that a student from one major would "fit" in the other. It is very robust when comparing a very large major to a small, specialized one.

# consolidated

```{r}
#| label: unified-stem-analysis
#| fig-cap: "Dynamic Curricular Similarity Analysis"
#| fig-height: 12
#| fig-width: 12

# ==========================================
# 1. CONFIGURATION (EDIT THIS SECTION)
#Options: "computing", "engineering", "sciences", "interdisciplinary"
analysis_mode <- "interdisciplinary" 

# Restrict to top N majors (ignored for "interdisciplinary" or fixed lists)
n_top_limit <- 25 
# ==========================================


# --- SELECTION LOGIC ---
target_cips <- NULL
chart_title_prefix <- ""

# Get all valid CIPS from dataset
all_cips <- unique(major_course_weights$cip6)

if (analysis_mode == "computing") {
  # DHS Computing List
  target_cips <- c(
    "110101", "110102", "110103", "110104", "110199", "110201", "110202", "110203",
    "110299", "110401", "110501", "110701", "110801", "110802", "110803",
    "110804", "110899", "110901", "111001", "111002", "111003", "111004", "111005",
    "111006", "111099", "140900", "140901", "140902", "140903", "140999", "151201",
    "151202", "151203", "151204", "151299", "151301"
  )
  chart_title_prefix <- "Computing"
  
} else if (analysis_mode == "engineering") {
  # Engineering (14) + Eng Tech (15)
  matches <- grep("^(14|15)", all_cips, value = TRUE)
  # Sort by size and take Top N
  pop_table <- major_sizes[cip6 %in% matches]
  setorder(pop_table, -total_grads)
  target_cips <- head(pop_table$cip6, n_top_limit)
  chart_title_prefix <- "Engineering"

} else if (analysis_mode == "sciences") {
  # Bio(26), Math(27), Phys(40), Inter(30), Ag(01), NatRes(03)
  matches <- grep("^(26|27|40|30|01|03)", all_cips, value = TRUE)
  pop_table <- major_sizes[cip6 %in% matches]
  setorder(pop_table, -total_grads)
  target_cips <- head(pop_table$cip6, n_top_limit)
  chart_title_prefix <- "Sciences"

} else if (analysis_mode == "interdisciplinary") {
  # One Representative from each branch
  target_cips <- c(
    "110701", # CS
    "141901", # Mech Eng
    "260101", # Bio
    "270101", # Math
    "400801", # Physics
    "400601"  # Geology
  )
  chart_title_prefix <- "STEM Branches"
}

# Filter to only CIPS that actually exist in the data
valid_cips <- intersect(target_cips, all_cips)

if(length(valid_cips) < 2) {
  stop(paste("Not enough majors found for mode:", analysis_mode))
}

# --- DATA SUBSETTING ---
subset_weights <- major_course_weights[cip6 %in% valid_cips]
subset_sets    <- major_course_sets[cip6 %in% valid_cips]


# --- METRIC CALCULATIONS ---

# 1. TF-IDF
dtm <- dcast(subset_weights, cip6 ~ course_id, value.var = "prob_taken", fill = 0)
mat <- as.matrix(dtm[, -1])
idf <- log(nrow(mat) / (colSums(mat > 0) + 1))
tfidf_mat <- t(t(mat) * idf)
sim_tfidf <- lsa::cosine(t(tfidf_mat)) * 100
rownames(sim_tfidf) <- dtm$cip6; colnames(sim_tfidf) <- dtm$cip6

# 2. Max-Scaled Cosine
dtm_max <- dcast(subset_weights, cip6 ~ course_id, value.var = "max_scaled", fill = 0)
sim_max <- lsa::cosine(t(as.matrix(dtm_max[, -1]))) * 100
rownames(sim_max) <- dtm_max$cip6; colnames(sim_max) <- dtm_max$cip6

# 3. PMI Similarity
total_n <- sum(subset_weights$student_count)
p_major <- subset_weights[, .(p_m = sum(student_count)/total_n), by = cip6]
p_course <- subset_weights[, .(p_c = sum(student_count)/total_n), by = course_id]
pmi_dt <- merge(subset_weights, p_major, by = "cip6")
pmi_dt <- merge(pmi_dt, p_course, by = "course_id")
pmi_dt[, pmi := log2((student_count/total_n) / (p_m * p_c))]
pmi_dt[pmi < 0, pmi := 0]
dtm_pmi <- dcast(pmi_dt, cip6 ~ course_id, value.var = "pmi", fill = 0)
sim_pmi <- lsa::cosine(t(as.matrix(dtm_pmi[, -1]))) * 100
rownames(sim_pmi) <- dtm_pmi$cip6; colnames(sim_pmi) <- dtm_pmi$cip6

# 4. Set-Based Metrics (Jaccard, Overlap, Dice, Kulczynski)
pairs <- CJ(cip_a = valid_cips, cip_b = valid_cips)

calc_metrics <- function(c1, c2) {
  set_a <- unlist(subset_sets[cip6 == c1, course_set])
  set_b <- unlist(subset_sets[cip6 == c2, course_set])
  if(length(set_a) == 0 | length(set_b) == 0) return(list(j=0, o=0, d=0, k=0))
  
  inter <- length(intersect(set_a, set_b))
  n_a <- length(set_a); n_b <- length(set_b)
  
  return(list(
    j = 100 * (inter / length(union(set_a, set_b))),
    o = 100 * (inter / min(n_a, n_b)),
    d = 100 * ((2 * inter) / (n_a + n_b)),
    k = 100 * (0.5 * ((inter / n_a) + (inter / n_b)))
  ))
}

res_list <- mapply(calc_metrics, pairs$cip_a, pairs$cip_b, SIMPLIFY = FALSE)
pairs[, jaccard := sapply(res_list, `[[`, "j")]
pairs[, overlap := sapply(res_list, `[[`, "o")]
pairs[, dice := sapply(res_list, `[[`, "d")]
pairs[, kulczynski := sapply(res_list, `[[`, "k")]

to_mat <- function(d, v) {
  m_d <- dcast(d, cip_a ~ cip_b, value.var = v)
  m <- as.matrix(m_d[, -1]); rownames(m) <- m_d$cip_a
  return(m)
}

sim_jac <- to_mat(pairs, "jaccard")
sim_ovl <- to_mat(pairs, "overlap")
sim_dice <- to_mat(pairs, "dice")
sim_kul <- to_mat(pairs, "kulczynski")


# --- PLOTTING ---
get_smart_label <- function(cip) {
  # Special overrides for Interdisciplinary mode for clarity
  if (analysis_mode == "interdisciplinary") {
    map <- c("110701"="CS", "141901"="MechEng", "260101"="Bio", "270101"="Math", "400801"="Physics", "400601"="Geology")
    if (cip %in% names(map)) return(map[cip])
  }
  
  # Standard Map Lookup
  if(exists("degree_names_map")) {
    local <- degree_names_map[cip6 == cip, degree]
    if(length(local) > 0 && !is.na(local)) return(local[1])
  }
  # Global STEM List Lookup
  if(exists("stem_cip_names") && cip %in% names(stem_cip_names)) {
    return(stem_cip_names[cip])
  }
  return(cip)
}

plot_hm <- function(mat, metric_name, pal) {
  if(is.null(mat) || nrow(mat) < 2) return(NULL)
  
  r_lab <- sapply(rownames(mat), get_smart_label)
  c_lab <- sapply(colnames(mat), get_smart_label)
  clean <- function(x) str_wrap(str_remove(x, "Bachelor of Science in "), 18)
  
  # Dynamic font size based on matrix size
  f_size <- if(nrow(mat) > 20) 7 else 9
  
  pheatmap(mat, 
           labels_row = clean(r_lab), labels_col = clean(c_lab),
           display_numbers = TRUE, number_format = "%.0f", 
           main = paste(chart_title_prefix, ":", metric_name), 
           fontsize = f_size, color = pal)
}

# Generate All Plots
plot_hm(sim_jac, "Jaccard (Curricular Width)", colorRampPalette(c("white", "#feedde", "#fd8d3c"))(100))
plot_hm(sim_tfidf, "TF-IDF (Specialized Identity)", colorRampPalette(c("white", "#deebf7", "#08306b"))(100))
plot_hm(sim_ovl, "Overlap (Subset Relations)", colorRampPalette(c("white", "#e5f5e0", "#31a354"))(100))
plot_hm(sim_pmi, "PMI (Signature Bonds)", colorRampPalette(c("white", "#fff7bc", "#d95f0e"))(100))
plot_hm(sim_max, "Max-Scaled (Core Structure)", colorRampPalette(c("white", "#f7fcf0", "#4eb3d3"))(100))
plot_hm(sim_dice, "Dice (Intersection Weighted)", colorRampPalette(c("white", "#f2f0f7", "#6a51a3"))(100))
plot_hm(sim_kul, "Kulczynski (Mutual Fit)", colorRampPalette(c("white", "#fff5f0", "#cb181d"))(100))

# --- EXPORT TO JSON ---

# 1. Create a readable label map for the JSON consumer
# This allows the JSON to keep concise keys (CIPs) but provides human names
cip_label_map <- sapply(valid_cips, function(x) {
  # Clean up the label similar to the plot function
  lbl <- get_smart_label(x)
  str_remove(lbl, "Bachelor of Science in ")
}, simplify = FALSE)

# 2. Course counts per CIP (unique courses taken by students in that major)
course_counts <- sapply(valid_cips, function(x) {
  s <- subset_sets[cip6 == x]$course_set
  if (length(s) > 0) length(s[[1]]) else 0L
}, simplify = FALSE)

# 3. Bundle all data
export_data <- list(
  metadata = list(
    mode = analysis_mode,
    top_limit = n_top_limit,
    generated_at = format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
    description = "Pairwise similarity matrices for STEM curricula."
  ),
  labels = cip_label_map,
  course_counts = course_counts,
  matrices = list(
    jaccard = sim_jac,
    tfidf = sim_tfidf,
    overlap = sim_ovl,
    pmi = sim_pmi,
    max_scaled = sim_max,
    dice = sim_dice,
    kulczynski = sim_kul
  )
)

# 4. Generate dynamic filename and save
json_filename <- paste0("similarity_metrics_", analysis_mode, ".json")
jsonlite::write_json(export_data, json_filename, pretty = TRUE, auto_unbox = TRUE)

message(paste("Successfully exported metric data to:", json_filename))
```

## saving figures

```{r}
#| label: unified-stem-analysis-figures
#| fig-cap: "Dynamic Curricular Similarity Analysis"
#| fig-height: 7
#| fig-width: 7

# ==========================================
# 1. CONFIGURATION (EDIT THIS SECTION)
# Options: "computing", "engineering", "sciences", "interdisciplinary"
analysis_mode <- "sciences" 

# Restrict to top N majors (ignored for "interdisciplinary" or fixed lists)
n_top_limit <- 10

# Visualization Settings
heatmap_font_size_normal <- 12  # For smaller matrices (<20 rows)
heatmap_font_size_dense  <- 12   # For larger matrices (>=20 rows)
heatmap_axis_font_size   <- 12   # Font size for row/column labels (set to NULL to use base size)
# ==========================================


# --- HELPER FUNCTIONS ---
get_smart_label <- function(cip) {
  # Special overrides for Interdisciplinary mode for clarity
  if (analysis_mode == "interdisciplinary") {
    map <- c("110701"="CS", "141901"="Mech Eng", "260101"="Bio", "270101"="Math", "400801"="Physics", "400601"="Geology")
    if (cip %in% names(map)) return(map[cip])
  }
  
  # Standard Map Lookup
  if(exists("degree_names_map")) {
    local <- degree_names_map[cip6 == cip, degree]
    if(length(local) > 0 && !is.na(local)) return(local[1])
  }
  # Global STEM List Lookup
  if(exists("stem_cip_names") && cip %in% names(stem_cip_names)) {
    return(stem_cip_names[cip])
  }
  return(cip)
}


# --- SELECTION LOGIC ---
target_cips <- NULL
chart_title_prefix <- ""

# Get all valid CIPS from dataset
all_cips <- unique(major_course_weights$cip6)

if (analysis_mode == "computing") {
  # DHS Computing List
  target_cips <- c(
    "110101", "110102", "110103", "110104", "110199", "110201", "110202", "110203",
    "110299", "110401", "110501", "110701", "110801", "110802", "110803",
    "110804", "110899", "110901", "111001", "111002", "111003", "111004", "111005",
    "111006", "111099", "140900", "140901", "140902", "140903", "140999", "151201",
    "151202", "151203", "151204", "151299", "151301"
  )
  chart_title_prefix <- "Computing"
  
} else if (analysis_mode == "engineering") {
  # Engineering (14) + Eng Tech (15)
  matches <- grep("^(14|15)", all_cips, value = TRUE)
  # Sort by size and take Top N
  pop_table <- major_sizes[cip6 %in% matches]
  setorder(pop_table, -total_grads)
  target_cips <- head(pop_table$cip6, n_top_limit)
  chart_title_prefix <- "Engineering"

} else if (analysis_mode == "sciences") {
  # Bio(26), Math(27), Phys(40), Inter(30), Ag(01), NatRes(03)
  matches <- grep("^(26|27|40|30|01|03)", all_cips, value = TRUE)
  # Sort by size and take Top N
  pop_table <- major_sizes[cip6 %in% matches]
  setorder(pop_table, -total_grads)
  target_cips <- head(pop_table$cip6, n_top_limit)
  chart_title_prefix <- "Sciences"

} else if (analysis_mode == "interdisciplinary") {
  # One Representative from each branch
  target_cips <- c(
    "110701", # CS
    "141901", # Mech Eng
    "260101", # Bio
    "270101", # Math
    "400801", # Physics
    "400601"  # Geology
  )
  chart_title_prefix <- "STEM Branches"
}

# 1. Intersect with data existence
valid_cips <- intersect(target_cips, all_cips)

# 2. Filter to only majors that have resolved names (exclude raw CIP codes)
# We use get_smart_label; if it returns the input CIP, no name was found.
valid_cips <- valid_cips[sapply(valid_cips, function(x) get_smart_label(x) != x)]

if(length(valid_cips) < 2) {
  stop(paste("Not enough NAMED majors found for mode:", analysis_mode))
}

# --- DATA SUBSETTING ---
subset_weights <- major_course_weights[cip6 %in% valid_cips]
subset_sets    <- major_course_sets[cip6 %in% valid_cips]


# --- METRIC CALCULATIONS ---

# 1. TF-IDF
dtm <- dcast(subset_weights, cip6 ~ course_id, value.var = "prob_taken", fill = 0)
mat <- as.matrix(dtm[, -1])
idf <- log(nrow(mat) / (colSums(mat > 0) + 1))
tfidf_mat <- t(t(mat) * idf)
sim_tfidf <- lsa::cosine(t(tfidf_mat)) * 100
rownames(sim_tfidf) <- dtm$cip6; colnames(sim_tfidf) <- dtm$cip6

# 2. Max-Scaled Cosine
dtm_max <- dcast(subset_weights, cip6 ~ course_id, value.var = "max_scaled", fill = 0)
sim_max <- lsa::cosine(t(as.matrix(dtm_max[, -1]))) * 100
rownames(sim_max) <- dtm_max$cip6; colnames(sim_max) <- dtm_max$cip6

# 3. PMI Similarity
total_n <- sum(subset_weights$student_count)
p_major <- subset_weights[, .(p_m = sum(student_count)/total_n), by = cip6]
p_course <- subset_weights[, .(p_c = sum(student_count)/total_n), by = course_id]
pmi_dt <- merge(subset_weights, p_major, by = "cip6")
pmi_dt <- merge(pmi_dt, p_course, by = "course_id")
pmi_dt[, pmi := log2((student_count/total_n) / (p_m * p_c))]
pmi_dt[pmi < 0, pmi := 0]
dtm_pmi <- dcast(pmi_dt, cip6 ~ course_id, value.var = "pmi", fill = 0)
sim_pmi <- lsa::cosine(t(as.matrix(dtm_pmi[, -1]))) * 100
rownames(sim_pmi) <- dtm_pmi$cip6; colnames(sim_pmi) <- dtm_pmi$cip6

# 4. Set-Based Metrics (Jaccard, Overlap, Dice, Kulczynski)
pairs <- CJ(cip_a = valid_cips, cip_b = valid_cips)

calc_metrics <- function(c1, c2) {
  set_a <- unlist(subset_sets[cip6 == c1, course_set])
  set_b <- unlist(subset_sets[cip6 == c2, course_set])
  if(length(set_a) == 0 | length(set_b) == 0) return(list(j=0, o=0, d=0, k=0))
  
  inter <- length(intersect(set_a, set_b))
  n_a <- length(set_a); n_b <- length(set_b)
  
  return(list(
    j = 100 * (inter / length(union(set_a, set_b))),
    o = 100 * (inter / min(n_a, n_b)),
    d = 100 * ((2 * inter) / (n_a + n_b)),
    k = 100 * (0.5 * ((inter / n_a) + (inter / n_b)))
  ))
}

res_list <- mapply(calc_metrics, pairs$cip_a, pairs$cip_b, SIMPLIFY = FALSE)
pairs[, jaccard := sapply(res_list, `[[`, "j")]
pairs[, overlap := sapply(res_list, `[[`, "o")]
pairs[, dice := sapply(res_list, `[[`, "d")]
pairs[, kulczynski := sapply(res_list, `[[`, "k")]

to_mat <- function(d, v) {
  m_d <- dcast(d, cip_a ~ cip_b, value.var = v)
  m <- as.matrix(m_d[, -1]); rownames(m) <- m_d$cip_a
  return(m)
}

sim_jac <- to_mat(pairs, "jaccard")
sim_ovl <- to_mat(pairs, "overlap")
sim_dice <- to_mat(pairs, "dice")
sim_kul <- to_mat(pairs, "kulczynski")


# --- PLOTTING ---
plot_hm <- function(mat, metric_name, pal, file_tag) {
  if(is.null(mat) || nrow(mat) < 2) return(NULL)
  
  r_lab <- sapply(rownames(mat), get_smart_label)
  c_lab <- sapply(colnames(mat), get_smart_label)
  
  # Clean and abbreviate labels
  clean <- function(x) {
    x <- str_remove(x, "Bachelor of Science in ")
    x <- str_remove(x, "Bachelor Arts in ")
    x <- str_replace_all(x, "Engineering", "Eng")
    x <- str_replace_all(x, "Technology", "Tech")
    x <- str_replace_all(x, "Environmental", "Env")
    x <- str_replace_all(x, "Science", "Sci")
    x <- str_replace_all(x, "Information", "Info")
    x <- str_replace_all(x, "Management", "Mgmt")
    x <- str_replace_all(x, "Systems", "Sys")
    x <- str_replace_all(x, "Agricultural", "Ag")
    str_wrap(x, 18)
  }
  
  # --- CHANGED: IEEE Single Column Optimization ---
  # Standard width is ~7 inches.
  # Font sizes controlled by configuration variables at top
  f_size <- if(nrow(mat) > 20) heatmap_font_size_dense else heatmap_font_size_normal
  
  # Use custom axis font size if specified, otherwise default to base font size
  axis_f_size <- if(!is.null(heatmap_axis_font_size)) heatmap_axis_font_size else f_size
  
  # 1. Save to File
  # Constructs filename like: heatmap_interdisciplinary_top25_jaccard.png
  fname <- paste0("heatmap_", analysis_mode, "_top", n_top_limit, "_", file_tag, ".png")
  
  pheatmap(mat, 
           labels_row = clean(r_lab), labels_col = clean(c_lab),
           display_numbers = TRUE, number_format = "%.0f", 
           main = paste(chart_title_prefix, ":", metric_name), 
           fontsize = f_size, 
           fontsize_number = f_size * 0.9, 
           fontsize_row = axis_f_size,
           fontsize_col = axis_f_size,
           color = pal,
           filename = fname,
           width = 7, height = 7) # 7x7 fits IEEE single col well

  # 2. Display in Output
  pheatmap(mat, 
           labels_row = clean(r_lab), labels_col = clean(c_lab),
           display_numbers = TRUE, number_format = "%.0f", 
           main = paste(chart_title_prefix, ":", metric_name), 
           fontsize = f_size, 
           fontsize_number = f_size * 0.9, 
           fontsize_row = axis_f_size,
           fontsize_col = axis_f_size,
           color = pal)
}

# Generate All Plots
plot_hm(sim_jac, "Jaccard\n(Curricular Width)", colorRampPalette(c("white", "#feedde", "#fd8d3c"))(100), "jaccard")
plot_hm(sim_tfidf, "TF-IDF\n(Specialized Identity)", colorRampPalette(c("white", "#deebf7", "#08306b"))(100), "tfidf")
plot_hm(sim_ovl, "Overlap\n(Subset Relations)", colorRampPalette(c("white", "#e5f5e0", "#31a354"))(100), "overlap")
plot_hm(sim_pmi, "PMI\n(Signature Bonds)", colorRampPalette(c("white", "#fff7bc", "#d95f0e"))(100), "pmi")
plot_hm(sim_max, "Max-Scaled\n(Core Structure)", colorRampPalette(c("white", "#f7fcf0", "#4eb3d3"))(100), "max_scaled")
plot_hm(sim_dice, "Dice\n(Intersection Weighted)", colorRampPalette(c("white", "#f2f0f7", "#6a51a3"))(100), "dice")
plot_hm(sim_kul, "Kulczynski\n(Mutual Fit)", colorRampPalette(c("white", "#fff5f0", "#cb181d"))(100), "kulczynski")

# --- EXPORT TO JSON ---

# 1. Create a readable label map for the JSON consumer
cip_label_map <- sapply(valid_cips, function(x) {
  lbl <- get_smart_label(x)
  str_remove(lbl, "Bachelor of Science in ")
}, simplify = FALSE)

# 2. Course counts per CIP
course_counts <- sapply(valid_cips, function(x) {
  s <- subset_sets[cip6 == x]$course_set
  if (length(s) > 0) length(s[[1]]) else 0L
}, simplify = FALSE)

# 3. Bundle all data
export_data <- list(
  metadata = list(
    mode = analysis_mode,
    top_limit = n_top_limit,
    generated_at = format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
    description = "Pairwise similarity matrices for STEM curricula."
  ),
  labels = cip_label_map,
  course_counts = course_counts,
  matrices = list(
    jaccard = sim_jac,
    tfidf = sim_tfidf,
    overlap = sim_ovl,
    pmi = sim_pmi,
    max_scaled = sim_max,
    dice = sim_dice,
    kulczynski = sim_kul
  )
)

# 4. Generate dynamic filename and save
json_filename <- paste0("similarity_metrics_", analysis_mode, "_top", n_top_limit, ".json")
jsonlite::write_json(export_data, json_filename, pretty = TRUE, auto_unbox = TRUE)

message(paste("Successfully exported metric data to:", json_filename))
```